Cooling Requirements for the Vertical Shear Instability in Protoplanetary Disks

### 2015-04-23 numerical.tex
notation change from epsilon to h

Unless otherwise stated, our fiducial disk parameters are 
$(\gamma, \Gamma) = (1.4, 1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.05)$, and 
vertical domain size $\zmax=5H$. This setup is effectively vertically
isothermal since $c_s^2(\zmax) \simeq 0.9c_s^2(0)$. In Fig. \ref{omega_z}
we plot the corresponding equilibrium rotation profile, which shows
that $\Omega$ decreases slightly away from the midplane.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/omega2} 
  \caption{Equilibrium rotation profile $\Omega(z)$,
    normalized by its mid-plane value, for  the fiducial disk model with $\Gamma=1.011$
    and $(p,q,\epsilon)=(-1.5,-1,0.05)$. 
    \label{omega_z} 
  }
\end{figure}

@@

% Fig. \ref{iso_eigen_kx} compares the numerically obtained eigenvalues 
% for the fundamental mode and that calculated from 
% Eq. \ref{simple_growth}. We find good agreement in the
% wave frequency $\omega$ at all $\khat$ and in growth rates $\nu$ for
% $\khat\gtrsim 10$. Growth rates are over-estimated by our analysis but 
% are $O(\epsilon\Omega_k)$, which is consistent with the discussion in
% Appendix \ref{max_growth1}. The match in $\nu$ worsens for 
% decreasing $\khat$ due to the neglect of global radial structure in
% the analysis. Nevertheless, given the number of approximations made
% in our analysis, the agreement is satisfactory.

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
%   0cm]{figures/compare_eigen_imag_iso} 
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
%   1cm]{figures/compare_eigen_real_iso}
%   \caption{Growth rate (top) and real frequency (bottom) of the
%     fundamental VSI mode in the fiducial disk model with $\beta =
%     10^{-3}$. 
%     % a disk with $(\gamma,
%     % \Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
%     % $\beta=10^{-3}$. 
%     The solid (dotted) line is obtained numerically
%     (analytically).  
%     \label{iso_eigen_kx} 
%   }
% \end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx10_analytic.ps}
  \caption{Unstable modes with $\khat=10$ in the fiducial disk model
    with $\beta=10^{-3}$. %  a disk with 
    % $(\gamma,\Gamma)=(1.4,1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.1)$ and
    % $\beta=10^{-3}$. 
    The fiducial boundary condition is a free surface ($\Delta P=0$)
    with vertical domain size $\zmax=5H$  (black dots).  Other
    cases correspond to different boundary conditions applied at
    different heights. The fundamental mode is marked by `f'. Lines
    are computed from the analytic dispersion relation 
    with thermal relaxation Eq. \ref{relax_disp} (dotted), and that for
    isothermal perturbations, Eq. \ref{simple_growth} (dashed). 
    \label{compare_modes_iso_kx10} 
  }
\end{figure}

@@

We further  test the dependence of $\beta_\mathrm{crit}$ on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a  reference
point and vary $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and
$\epsilon\in[0.02,0.1]$ separately. The numerically obtained
$\beta_\mathrm{crit}$ are shown in Fig. \ref{bcrit_compare} in
comparison with Eq. \ref{iso_vsi_cond}, which generally agree. This
result, together with Fig. \ref{bcrit_compare1}, shows that  
$\beta \lesssim \beta_\mathrm{crit}$ is a practical
criteria for the VSI to operate effectively in stably stratified,
vertically isothermal disks.

% The agreement improves with decreasing  $|q|$,
% $\epsilon$ and increasing $\gamma$, i.e. weaker instability. There is
% a noticeable difference as $\gamma\to1$ because the derivation of
% $\beta_\mathrm{crit}$ assumed $\gamma>1$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0cm
  0cm]{figures/bcrit_compare_q.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.8cm]{figures/bcrit_compare_g.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.8cm]{figures/bcrit_compare_e.ps} 
  \caption{Dependence of the upper limit to the thermal relaxation timescale
    $\beta_\mathrm{crit}$ for the fundamental VSI mode on disk
    parameters. The fiducial setup is $(\gamma, \Gamma= (1.4, 1.011)$
    and $(p,q,\epsilon)=(-1.5,-1,0.05)$. Top: varying 
    $q\in[-1.2,-0.2]$; middle: varying $\gamma\in[1.2,2.0]$; bottom:
    varying $\epsilon\in[0.02,0.1]$. The perturbation wavenumber is
    $\khat=10$.  
    \label{bcrit_compare}}  
\end{figure}

@@

Nevertheless, here we briefly consider a vertically non-isothermal
disk model with $(p,q,\epsilon)=(0,-1,0.05)$ and $\Gamma=1.0837$. This
value of $\Gamma$ is chosen so that the standard vertical domain size $\zmax =
5H$ effectively corresponds to the disk surface ($\zmax\simeq
H_s$). The disk temperature decreases rapidly with height such that 
$c_s^2(\zmax)\simeq 10^{-4}c_s^2(0)$.

@@

Finally, in Fig. \ref{bcrit_compare2} we plot the maximum VSI growth
rates in the vertically non-isothermal disk model. Based on 
Eq. \ref{iso_vsi_cond}, we conjecture the critical 
thermal timescale in the vertically non-isothermal disk is 
given by $\epsilon|s|/(\gamma - \Gamma)$. This 
evaluates to 0.158, which is close to the cut-off thermal timescale
for $\khat\in[5,30]$. Fig. \ref{bcrit_compare2} is qualitatively 
similar to the fiducial disk model in
Fig. \ref{bcrit_compare1}. Importantly, for $\khat=O(10)$ there is a
precise cut-off thermal timescale, the value of which in fact provides a  
characteristic upper limit for other $\khat$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gcorr_compare_vnoniso_maxrate} 
  \caption{Maximum VSI growth rates in the vertically non-isothermal disk
    model as a function of the thermal relaxation timescale
    $\beta$. The vertical line is a conjectured 
    critical thermal timescale given by
    $\epsilon|s|/(\gamma-\Gamma)=0.158$.  
    \label{bcrit_compare2}}   
\end{figure}

  % We briefly consider vertically non-isothermal disks with 
% $\Gamma=1.4$ and $(p,q,\epsilon)=(0,-1,0.05)$ as simulated in
% \cite{nelson13}. In this case we set $\zmax=0.99H_s$.

@@

% For $\gamma>\Gamma$, i.e. stably stratified disks,
% Fig. \ref{gcorr_compare_vnoniso} shows that introducing finite thermal
% relaxation rapidly stabilizes the disk, similar to that observed for
% nearly vertically isothermal disks (Fig. \ref{bcrit_compare1}). For
% $\gamma=1.7,\,2.0,\,2.5$, growth rates reach zero at
% $\beta\simeq0.17,\,0.083,\,0.045$, respectively. Interestingly, these
% values are equal to $\epsilon|q|/(\gamma-\Gamma)$. This suggests that
% the critical thermal relaxation timescale for vertically
% non-isothermal disks can also be estimated by Eq. \ref{iso_vsi_cond}
% but with $\gamma-1$ replaced by $\gamma-\Gamma$.

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
%   0cm]{figures/gcorr_compare_vnoniso2}
%   \caption{Growth rate of the fundamental VSI mode as a function of
%     the thermal relaxation timescale $\beta$, in vertically
%     non-isothermal disks with $\Gamma=1.4$ and
%     $\gamma\in[1.4,2.5]$. The disk is neutrally
%     stratified for $\gamma=1.4$ and stably stratified for
%     $\gamma>1.4$. Other disk parameters are
%     $(p,q,\epsilon)=(0,-1,0.05)$ and the perturbation wavenumber is
%     $\khat=30$.   
%     \label{gcorr_compare_vnoniso}}
% \end{figure}

>>>>

Unless otherwise stated, our fiducial disk parameters are 
$(\gamma, \Gamma) = (1.4, 1.011)$, $(p,q, h)=(-1.5,-1,0.05)$, and 
vertical domain size $\zmax=5H$. This setup is effectively vertically
isothermal since $c_s^2(\zmax) \simeq 0.9c_s^2(0)$. In Fig. \ref{omega_z}
we plot the corresponding equilibrium rotation profile, which shows
that $\Omega$ decreases slightly away from the midplane.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/omega2} 
  \caption{Equilibrium rotation profile $\Omega(z)$,
    normalized by its mid-plane value, for  the fiducial disk model with $\Gamma=1.011$
    and $(p,q, h)=(-1.5,-1,0.05)$. 
    \label{omega_z} 
  }
\end{figure}

@@

% Fig. \ref{iso_eigen_kx} compares the numerically obtained eigenvalues 
% for the fundamental mode and that calculated from 
% Eq. \ref{simple_growth}. We find good agreement in the
% wave frequency $\omega$ at all $\khat$ and in growth rates $\nu$ for
% $\khat\gtrsim 10$. Growth rates are over-estimated by our analysis but 
% are $O( h\Omega_k)$, which is consistent with the discussion in
% Appendix \ref{max_growth1}. The match in $\nu$ worsens for 
% decreasing $\khat$ due to the neglect of global radial structure in
% the analysis. Nevertheless, given the number of approximations made
% in our analysis, the agreement is satisfactory.

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
%   0cm]{figures/compare_eigen_imag_iso} 
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
%   1cm]{figures/compare_eigen_real_iso}
%   \caption{Growth rate (top) and real frequency (bottom) of the
%     fundamental VSI mode in the fiducial disk model with $\beta =
%     10^{-3}$. 
%     % a disk with $(\gamma,
%     % \Gamma)=(1.4,1.011)$, $(p,q, h)=(-1.5,-1,0.1)$ and
%     % $\beta=10^{-3}$. 
%     The solid (dotted) line is obtained numerically
%     (analytically).  
%     \label{iso_eigen_kx} 
%   }
% \end{figure}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/compare_modes_iso_kx10_analytic.ps}
  \caption{Unstable modes with $\khat=10$ in the fiducial disk model
    with $\beta=10^{-3}$. %  a disk with 
    % $(\gamma,\Gamma)=(1.4,1.011)$, $(p,q, h)=(-1.5,-1,0.1)$ and
    % $\beta=10^{-3}$. 
    The fiducial boundary condition is a free surface ($\Delta P=0$)
    with vertical domain size $\zmax=5H$  (black dots).  Other
    cases correspond to different boundary conditions applied at
    different heights. The fundamental mode is marked by `f'. Lines
    are computed from the analytic dispersion relation 
    with thermal relaxation Eq. \ref{relax_disp} (dotted), and that for
    isothermal perturbations, Eq. \ref{simple_growth} (dashed). 
    \label{compare_modes_iso_kx10} 
  }
\end{figure}

@@

We further  test the dependence of $\beta_\mathrm{crit}$ on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a  reference
point and vary $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and
$ h\in[0.02,0.1]$ separately. The numerically obtained
$\beta_\mathrm{crit}$ are shown in Fig. \ref{bcrit_compare} in
comparison with Eq. \ref{iso_vsi_cond}, which generally agree. This
result, together with Fig. \ref{bcrit_compare1}, shows that  
$\beta \lesssim \beta_\mathrm{crit}$ is a practical
criteria for the VSI to operate effectively in stably stratified,
vertically isothermal disks.

% The agreement improves with decreasing  $|q|$,
% $ h$ and increasing $\gamma$, i.e. weaker instability. There is
% a noticeable difference as $\gamma\to1$ because the derivation of
% $\beta_\mathrm{crit}$ assumed $\gamma>1$.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.cm 0cm
  0cm]{figures/bcrit_compare_q.ps} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.8cm]{figures/bcrit_compare_g.ps}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0.0cm 0cm
  0.8cm]{figures/bcrit_compare_e.ps} 
  \caption{Dependence of the upper limit to the thermal relaxation timescale
    $\beta_\mathrm{crit}$ for the fundamental VSI mode on disk
    parameters. The fiducial setup is $(\gamma, \Gamma= (1.4, 1.011)$
    and $(p,q, h)=(-1.5,-1,0.05)$. Top: varying 
    $q\in[-1.2,-0.2]$; middle: varying $\gamma\in[1.2,2.0]$; bottom:
    varying $ h\in[0.02,0.1]$. The perturbation wavenumber is
    $\khat=10$.  
    \label{bcrit_compare}}  
\end{figure}

@@

Nevertheless, here we briefly consider a vertically non-isothermal
disk model with $(p,q, h)=(0,-1,0.05)$ and $\Gamma=1.0837$. This
value of $\Gamma$ is chosen so that the standard vertical domain size $\zmax =
5H$ effectively corresponds to the disk surface ($\zmax\simeq
H_s$). The disk temperature decreases rapidly with height such that 
$c_s^2(\zmax)\simeq 10^{-4}c_s^2(0)$.

@@

Finally, in Fig. \ref{bcrit_compare2} we plot the maximum VSI growth
rates in the vertically non-isothermal disk model. Based on 
Eq. \ref{iso_vsi_cond}, we conjecture the critical 
thermal timescale in the vertically non-isothermal disk is 
given by $ h|s|/(\gamma - \Gamma)$. This 
evaluates to 0.158, which is close to the cut-off thermal timescale
for $\khat\in[5,30]$. Fig. \ref{bcrit_compare2} is qualitatively 
similar to the fiducial disk model in
Fig. \ref{bcrit_compare1}. Importantly, for $\khat=O(10)$ there is a
precise cut-off thermal timescale, the value of which in fact provides a  
characteristic upper limit for other $\khat$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gcorr_compare_vnoniso_maxrate} 
  \caption{Maximum VSI growth rates in the vertically non-isothermal disk
    model as a function of the thermal relaxation timescale
    $\beta$. The vertical line is a conjectured 
    critical thermal timescale given by
    $ h|s|/(\gamma-\Gamma)=0.158$.  
    \label{bcrit_compare2}}   
\end{figure}

  % We briefly consider vertically non-isothermal disks with 
% $\Gamma=1.4$ and $(p,q, h)=(0,-1,0.05)$ as simulated in
% \cite{nelson13}. In this case we set $\zmax=0.99H_s$.

@@

% For $\gamma>\Gamma$, i.e. stably stratified disks,
% Fig. \ref{gcorr_compare_vnoniso} shows that introducing finite thermal
% relaxation rapidly stabilizes the disk, similar to that observed for
% nearly vertically isothermal disks (Fig. \ref{bcrit_compare1}). For
% $\gamma=1.7,\,2.0,\,2.5$, growth rates reach zero at
% $\beta\simeq0.17,\,0.083,\,0.045$, respectively. Interestingly, these
% values are equal to $ h|q|/(\gamma-\Gamma)$. This suggests that
% the critical thermal relaxation timescale for vertically
% non-isothermal disks can also be estimated by Eq. \ref{iso_vsi_cond}
% but with $\gamma-1$ replaced by $\gamma-\Gamma$.

% \begin{figure}
%   \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
%   0cm]{figures/gcorr_compare_vnoniso2}
%   \caption{Growth rate of the fundamental VSI mode as a function of
%     the thermal relaxation timescale $\beta$, in vertically
%     non-isothermal disks with $\Gamma=1.4$ and
%     $\gamma\in[1.4,2.5]$. The disk is neutrally
%     stratified for $\gamma=1.4$ and stably stratified for
%     $\gamma>1.4$. Other disk parameters are
%     $(p,q, h)=(0,-1,0.05)$ and the perturbation wavenumber is
%     $\khat=30$.   
%     \label{gcorr_compare_vnoniso}}
% \end{figure}

### 2015-04-23 summary.tex
notation change from epsilon to h

For a vertically isothermal, sub-adiabatically stratified disk we
determined the thermal timescale   
\begin{align*}
  t_\mathrm{crit} = \frac{\epsilon |q|}{\gamma-1}\Omega_k^{-1}, 
\end{align*}
below which the VSI operates effectively. While $t_\mathrm{crit}$ was
derived for the lowest order VSI mode --- the `fundamental  
corrugation mode' --- we find from numerical calculations that
$t_\mathrm{crit}$ can be used to characterize the general
stabilization of the VSI by a finite thermal timescale. %That is, all
%VSI growth rates are reduced significantly as $\beta$ is increased
%from 
%zero to $\beta_\mathrm{crit}$.    
We have therefore provided a quantitative measure for the thermal
timescale requirement for the VSI in astrophysical 
disks. The thermal timescale must be shorter than the disk
dynamical timescale by $O(\epsilon)$, which is
quite short for thin disks ($\epsilon\ll 1$), because vertical shear is weak.

@@

The convective overstability corresponds to slowly growing epicyclic motions.
It is most effective for $t_c\Omega \sim 1$ and disturbances with
$|k_x|\ll |k_z|$, where $k_z$ is a vertical wavenumber. These
conditions are mutually exclusive with the VSI.  
The maximum growth rate for convective overstability is given by 
\begin{align}
  \nu_\mathrm{conv,max} = -\frac{N^2_r}{4\Omega}\label{max_conv_gen}
\end{align}
\citep{lyra14}, where $N_r^2 = -(c_p\rho)^{-1}\p_rP\p_rS$ is the
square of the radial buoyancy frequency
(cf. Eq. \ref{nzsq_def}). Thus, growth requires $N_r^2<0$.  
%Evaluating Eq. \ref{max_conv_gen} at the midplane of our disk models,
%we find  
%\begin{align}
%  \nu_\mathrm{conv,max} =
%  -\frac{\epsilon^2\Omega_k}{4\gamma\Gamma}\left(p+q\right)
%  \left[\left(\gamma-1\right)p-q\right]. \label{max_conv}
%\end{align}
In fact, for the MMSN considered in this work, convective
overstability does not apply because 
$N_r^2>0$.

>>>>

For a vertically isothermal, sub-adiabatically stratified disk we
determined the thermal timescale   
\begin{align*}
  t_\mathrm{crit} = \frac{ h |q|}{\gamma-1}\Omega_k^{-1}, 
\end{align*}
below which the VSI operates effectively. While $t_\mathrm{crit}$ was
derived for the lowest order VSI mode --- the `fundamental  
corrugation mode' --- we find from numerical calculations that
$t_\mathrm{crit}$ can be used to characterize the general
stabilization of the VSI by a finite thermal timescale. %That is, all
%VSI growth rates are reduced significantly as $\beta$ is increased
%from 
%zero to $\beta_\mathrm{crit}$.    
We have therefore provided a quantitative measure for the thermal
timescale requirement for the VSI in astrophysical 
disks. The thermal timescale must be shorter than the disk
dynamical timescale by $O( h)$, which is
quite short for thin disks ($ h\ll 1$), because vertical shear is weak.

@@

The convective overstability corresponds to slowly growing epicyclic motions.
It is most effective for $t_c\Omega \sim 1$ and disturbances with
$|k_x|\ll |k_z|$, where $k_z$ is a vertical wavenumber. These
conditions are mutually exclusive with the VSI.  
The maximum growth rate for convective overstability is given by 
\begin{align}
  \nu_\mathrm{conv,max} = -\frac{N^2_r}{4\Omega}\label{max_conv_gen}
\end{align}
\citep{lyra14}, where $N_r^2 = -(c_p\rho)^{-1}\p_rP\p_rS$ is the
square of the radial buoyancy frequency
(cf. Eq. \ref{nzsq_def}). Thus, growth requires $N_r^2<0$.  
%Evaluating Eq. \ref{max_conv_gen} at the midplane of our disk models,
%we find  
%\begin{align}
%  \nu_\mathrm{conv,max} =
%  -\frac{ h^2\Omega_k}{4\gamma\Gamma}\left(p+q\right)
%  \left[\left(\gamma-1\right)p-q\right]. \label{max_conv}
%\end{align}
In fact, for the MMSN considered in this work, convective
overstability does not apply because 
$N_r^2>0$.
